---
title: "Moving Weighted Average - Test on SCEVAN & ASCAT Segmentation Files"
output: none
---

# install dependencies
```{r, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

if (!require("IRanges", quietly = TRUE))
    BiocManager::install("IRanges")
```

# import dependencies
```{r}
library(IRanges)
library(dplyr)
```


# define envrionment variable (won't be used in this test, just writing it here so it's available for copy-paste into the version used to analyze all our data)
```{r}
# index <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))
index <- 1
```


# load input mapping file
```{r}
map <- read.csv("/scratch/dgladish/benchmarking_spatial_scale/mapping_files/alg_out_paths.csv")
print(paste("Analysis of sample", map$sample_id[index]))
```



# import data
```{r}
ascat_segs <- read.delim(map$ascat_seg_path[index], stringsAsFactors = FALSE)
scevan_segs <- read.delim(map$scevan_seg_path[index], stringsAsFactors = FALSE)
```

# import Ted's chromosome coordinates
```{r}
path_chrom_lengths <- "/scratch/dgladish/benchmarking_spatial_scale/chrom_lengths/HG38.txt"
chrom_lengths <- read.delim(path_chrom_lengths, stringsAsFactors = FALSE, header=F)
chrom_lengths[, 3] <- sub("chr", "", chrom_lengths[, 1])

chrom_lengths[, 3] <- sub("X", "23", chrom_lengths[, 3])
chrom_lengths[, 3] <- sub("Y", "24", chrom_lengths[, 3])
```


# make copy number values relative to 2
```{r}
# ASCAT
ascat_segs$Copy_Number <- ascat_segs$Copy_Number-2
print(ascat_segs$Copy_Number)
```




```{r}
# SCEVAN
scevan_segs$CN <- scevan_segs$CN-2
print(scevan_segs$CN)
```

# calculate window averages for ASCAT segments
## add in an additional chromosome column to the ASCAT segmentation file
```{r}
ascat_segs$Chromosome_numeric <- sub("chr", "", ascat_segs$Chromosome)
ascat_segs$Chromosome_numeric <- sub("X", "23", ascat_segs$Chromosome_numeric)
ascat_segs$Chromosome_numeric <- sub("Y", "24", ascat_segs$Chromosome_numeric)

ascat_segs$Chromosome_numeric <- as.numeric(ascat_segs$Chromosome_numeric)
```


## define window size
```{r}
window_size <- 1000000
```

## calculate the # of windows roughly that will be used for each chromosome (HG38 lengths, not ASCAT, so results might be a bit different, but should be identical or very close)
```{r}
windows_per_chrom <- ceiling(chrom_lengths[, 2] / window_size)
print(windows_per_chrom)
```
## calculate the total # of windows roughly that will be used in this analysis
```{r}
total_windows <- sum(windows_per_chrom)
print(total_windows)
```



## run the analysis
```{r}
ascat_WAs <- numeric() # vector will be filled by each iteration, first with the chromosome 1 windows, then chromosome 2, etc etc

for (i in 1:max(ascat_segs$Chromosome_numeric)) {
  #browser()
  ascat_segs_subset <- ascat_segs[which(ascat_segs$Chromosome_numeric==i), ] # create the subset of the data frame so that you are only analyzing 1 chromosome at a time
  
  seg_ranges <- IRanges(start = ascat_segs_subset$Start, end = ascat_segs_subset$End) 
  chrom_length <- chrom_lengths[which(chrom_lengths[, 3]==i), 2]
  
  # Set up fixed windows of 1Mb
  windows <- IRanges(start = seq(1, chrom_length, by = window_size), width = window_size)
  
  ascat_loop_WAs <-  sapply(seq_along(windows), function(j) {
    #browser()
    win <- windows[j]
    win_repeated <- rep(win, length(seg_ranges))
    overlaps <- pintersect(seg_ranges, win_repeated, resolve.empty = "start")
    overlap_widths <- width(overlaps)
    
    valid_idx <- which(overlap_widths > 0)
    if (length(valid_idx) == 0) return(NA)
    
    overlap_weights <- overlap_widths[valid_idx] / width(win)
    copy_vals <- ascat_segs_subset$Copy_Number[valid_idx]
    
    sum(copy_vals * overlap_weights)
    
    
  })
  names(ascat_loop_WAs) <- rep(ascat_segs_subset$Chromosome[1], times=length(ascat_loop_WAs))
  
  ascat_WAs <- c(ascat_WAs, ascat_loop_WAs)
}
```

# Looks good! Now calculate moving averages for the SCEVAN data!
## initialize an empty vector that will hold the output of the loop
```{r}
scevan_WAs <- numeric()

for (i in 1:max(scevan_segs$Chr)) {
  #browser()
  scevan_segs_subset <- scevan_segs[which(scevan_segs$Chr==i), ] # create the subset of the data frame so that you are only analyzing 1 chromosome at a time
  
  seg_ranges <- IRanges(start = scevan_segs_subset$Pos, end = scevan_segs_subset$End) # convert the data frame holding the subset of the segmentation data to an IRanges object so that the IRanges function pintersect() may be used on the data
  chrom_length <- chrom_lengths[which(chrom_lengths[, 3]==i), 2] # determine the end point of the chromosomes. This will be used to determine the number of windows that will be used in the analysis of the chromosome being analyzed in a given iteration.
  
  # Set up fixed windows of 1Mb
  windows <- IRanges(start = seq(1, chrom_length, by = window_size), width = window_size) 
  
  # for each chromosome, this analysis will be looped, shifting the window each time (i.e. chr1 is first analyzed   with 1-1,000,000, then 1,000,001-2,000,000, etc. then repeat for chromosome 2, then 3, etc.).
  scevan_loop_WAs <- sapply(seq_along(windows), function(j) { 
    #browser()
    win <- windows[j] # selects the window to use (e.g. 1-1,000,000)
    win_repeated <- rep(win, length(seg_ranges))
    overlaps <- pintersect(seg_ranges, win_repeated, resolve.empty = "start")
    overlap_widths <- width(overlaps)
    
    valid_idx <- which(overlap_widths > 0)
    if (length(valid_idx) == 0) return(NA) # 
    
    overlap_weights <- overlap_widths[valid_idx] / width(win)
    copy_vals <- scevan_segs_subset$CN[valid_idx]
    
    sum(copy_vals * overlap_weights)
    
    
  })
  #browser()
  current_chrom <- scevan_segs_subset$Chr[1]
  for_names_scevan_loop_WAs <- sub(current_chrom, paste(c("chr", current_chrom), collapse=""), current_chrom)
  names(scevan_loop_WAs) <- rep(for_names_scevan_loop_WAs, times=length(scevan_loop_WAs))
  
  scevan_WAs <- c(scevan_WAs, scevan_loop_WAs)
}
```

## do a quick investigation of the output so I understand what I am working with
### NAs
#### are there any NA values? if so, where are they localized/concentrated?
##### ASCAT
```{r}
if (any(is.na(ascat_WAs))) {
  print(paste("There are", sum(is.na(ascat_WAs)), "NAs in ascat_WAs! There was at least 1 window that did not have any segment overlap!"))
  
  # convert MA output vector to a list
  #browser()
  #original_order <- names(ascat_WAs)
  ascat_WAs_list <- split(ascat_WAs, names(ascat_WAs))
  na_positions_normalized_list <- lapply(ascat_WAs_list, function(x) {which(is.na(x)) / length(x)}) # extract locations of the NA values & normalize by chromosome length (indices closer to 0 will be front of the chromosome & those closer to 1 will be towards the end of the chromosome)
  na_positions_normalized_vec <- unlist(na_positions_normalized_list, use.names = TRUE) # create a named vector from the list containing the normalized NA locations
  #na_positions_normalized <- na_positions_normalized[original_order]
  
  # create a histogram visualizing the locations of the NA values
  hist(na_positions_normalized_vec, breaks = 10, main = "ASCAT - Empty Window Locations",
     xlab = "Normalized Window Position (Index in Vector / Vector Length)",
     col = "#00F5FF", border = "white")


  # bar plots
  ## number of windows / chromosome
  #browser()
  original_order <- unique(names(ascat_WAs))
  windows_per_chrom_ascat <- sapply(ascat_WAs_list, length)
  windows_per_chrom_ascat <- windows_per_chrom_ascat[original_order]
  
  barplot(windows_per_chrom_ascat, las=2, col = "skyblue", main = "ASCAT - Total Windows Per Chromosome", ylim = c(0, 250))
  
  ## number of empty windows / chromosome
  nas_per_chrs <- sapply(ascat_WAs_list, function(x) {sum(is.na(x))})
  nas_per_chrs <- nas_per_chrs[original_order]
  barplot(nas_per_chrs, las=2, col="blue", main = "ASCAT - Empty Windows Per Chromosome", ylim = c(0, 25))
  
  } else {
  print("There are no NAs in ascat_WAs! All windows had at least 1 segment span them.")
  }
```
##### SCEVAN 
```{r}
if (any(is.na(scevan_WAs))) {
  print(paste("There are", sum(is.na(scevan_WAs)), "NAs in scevan_WAs! There was at least 1 window that did not have any segment overlap!"))
  
  # histogram
  ## convert MA output vector to a list
  scevan_WAs_list <- split(scevan_WAs, names(scevan_WAs))
  
  scevan_ordered_names <- unique(names(scevan_WAs))
  scevan_WAs_list <- scevan_WAs_list[scevan_ordered_names]
  
  na_positions_normalized <- lapply(scevan_WAs_list, function(x) {which(is.na(x)) / length(x)}) # extract locations of the NA values & normalize by chromosome length (indices closer to 0 will be front of the chromosome & those closer to 1 will be towards the end of the chromosome)
  na_positions_normalized <- unlist(na_positions_normalized, use.names = TRUE) # create a named vector from the list containing the normalized NA locations
  
  ## create a histogram visualizing the locations of the NA values
  hist(na_positions_normalized, breaks = 10, main = "SCEVAN - Empty Window Locations",
     xlab = "Normalized Window Position (Index in Vector / Vector Length)",
     col = "deeppink", border = "white")
  
  # bar plots
  ## number of windows / chromosome
  windows_per_chrom_scevan <- sapply(scevan_WAs_list, length)
  windows_per_chrom_scevan <- windows_per_chrom_scevan[original_order]
  
  barplot(windows_per_chrom_scevan, las=2, col = "#FFE1FF", main = "SCEVAN - Number of Total Windows Per Chromosome", ylim=c(0, 250))
  
  ## number of empty windows / chromosome
  ## sum the number of NA windows for each chromosome
  nas_per_chrs <- sapply(scevan_WAs_list, function(x) {sum(is.na(x))})
  nas_per_chrs <- nas_per_chrs[original_order]
  
  ## create the bar plot showing the number of empty windows / chromosome
  barplot(nas_per_chrs, las=2, col = "red", main = "SCEVAN - Number of Empty Windows Per Chromosome", ylim=c(0, 25))
  
  } else {
    
  print("There are no NAs in scevan_WAs! All windows had at least 1 segment span them.")
    
  }
```
##### Also, create a bar plot that shows the # of windows that do not have windows in them for SCEVAN
```{r}
num_NAs <-c(sum(is.na(scevan_WAs)), sum(is.na(ascat_WAs)))
names(num_NAs) <- c("SCEVAN", "ASCAT")

barplot(num_NAs, main="Number of Empty Windows for Each Algorithm", col=c("#C1440E", "#3B6BA5"), ylim=c(0, 100))
```

# Add back 2 to each of the windows. The mathematical property 
```{r}
ascat_WAs <- ascat_WAs+2
scevan_WAs <- scevan_WAs+2
```
